A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-05-18 JNABsub20230518/ComparingNLStools.Rmd
May 18 fixes

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
date: "2023-5-16"
abstract: >
  Our Google Summer of Code project "Improvements to `nls()`" considered the 
  features and limitations of the R function `nls()` with the aim 
  of improving and rationalizing R tools for nonlinear regression.
  The rich features of `nls()` are weakened by several deficiencies and
  inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
  Further considerations are the usability and  maintainability of
  the code base that provides the functionality `nls()` claims to offer.
  Various packages, including our `nlsr`, provide alternative capabilities.
  We consider the differences in goals, approaches, and features of
  different tools for nonlinear least squares modeling in R. Discussion of 
  these matters is relevant to improving R generally as well as its 
  nonlinear estimation tools.

draft: true
author:
  # see ?rjournal_article for more information
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output:
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: ./ComparingNLStools.bib

@@

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

As mentioned, the output of `nls()` is an object of class "nls" which has a quite rich structure
described in the manual file or revealed by applying the `str()` function to the
result of `nls()`.

The complexity of this object is a challenge to users. Let us use the
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
actually access the data from this element, we need to use the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Bounds on parameters raise some interesting questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
errors nor their derived statistics when bounds are active.

@@

<!-- Ben Bolker mentioned a possibility of using a nonlinear model in  -->
<!-- ??I think?? nlme with some sort of linear model form of the parameters. -->
<!-- May need to mention. -->

<!-- ===```{r nlsindx1} -->
<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
<!-- ## contraction on 21 animals.  The observed variables are Strip -->
<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
<!-- ## (resulting length of muscle section). -->
<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE) -->
<!-- mm<-droplevels(mm) # remove unused levels after truncation -->
<!-- nlev <- nlevels(mm) -->
<!-- withAutoprint({ -->
<!--   ## The non linear model considered is -->
<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
<!--   ## where theta is constant. For now alpha and beta do NOT vary with Strip. -->
<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
<!--   nl <- nlevels(mm$Strip) -->
<!--   ## We first use the plinear algorithm to fit an overall model, -->
<!--   ## ignoring that alpha and beta might vary with Strip. -->
<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
<!--                 start = list(th = 1), algorithm = "plinear") -->
<!--   summary(musc.1) -->

<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
<!--   ## models to use the conventional algorithm to fit a model in which -->
<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
<!--   ## by the previously fitted model. -->
<!--   ## Note that with indexed parameters, the starting values must be -->
<!--   ## given in a list (with names): -->
<!--   b <- coef(musc.1) -->
<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
<!--   summary(musc.2) -->
<!-- }) -->
<!-- === ``` -->

@@

<!-- ## Should nls() be refactored? -->

<!-- When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted: -->

<!-- - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R  -->
<!--   that matches the features in base R or what is packaged in `nlspkg` as described  -->
<!--   in @PkgFromRbase22. Our intent in providing a purely R version of the code was -->
<!--   either to replace `nls()` or to serve as a reference version aiding maintenance  -->
<!--   of C, Fortran, or other components of the software. -->

<!-- - to streamline the overall `nls()` infrastructure. By this we mean a  -->
<!--   refactoring of the routines so they are better suited to maintenance of both -->
<!--   the existing `nls()` methods and features as well as new features we or others -->
<!--   would like to add. -->

<!-- - to explain what we do, either in comments, examples, or separate maintainer  -->
<!--   documentation. While R and its packages are generally very well documented for -->
<!--   usage, that documentation -->
<!--   rarely extends as far as it might for code maintenance. The paucity of such -->
<!--   documentation is exacerbated by the mixed R/C/Fortran code base, -->
<!--   where seemingly trivial differences in things like representation of numbers or -->
<!--   base index for arrays lead to traps for unwary programmers. -->

<!-- These goals echo themes in the recent discussion by @Vidoni21 and accompanying  -->
<!-- commentary. -->

<!-- We found the legacy code and examples prevented us achieving a revised `nls()`  -->
<!-- with all its current capabilities. Unfortunately, `nls()` is part of the base R -->
<!-- distribution,  -->
<!-- so users are drawn to using it even when it is less than ideal. A possible way forward -->
<!-- is to  -->

<!-- - use a packaged version of `nls()` to provide legacy features, with notification -->
<!--   to users that there are other tools available -->

<!-- - if maintainers are willing, to merge and rationalize the capabilities of  -->
<!--   `minpack.lm`, `nlsr` and `gslnls` -->

<!-- - to encourage the (separate) development of capabilities such as indexed -->
<!--   parameter models if these are in demand. -->

<!-- This direction does not, of course, resolve the inconsistencies in model  -->
<!-- specification in formulas for nonlinear models, which undoubtedly is an issue -->
<!-- for other modeling tools.   -->

<!-- A possible workaround to allow consistency for the `plinear` algorithm  -->
<!-- would be to specify the linear parameters when the `plinear` option is specified.  -->
<!-- For example, we could use `algorithm="plinear(Asym)"` while keeping the general  -->
<!-- model formula from the default specification. This would allow  -->
<!-- for the output of different `algorithm` options to be consistent. However, we  -->
<!-- have not yet tried to code such a change, and welcome discussion from those working -->
<!-- with other packages using model formulas. -->

<!-- A further complication of model specification in R is that some formulas require -->
<!-- the user to surround a term with `I()` to inhibit interpretation of arithmetic  -->
<!-- operators. We would prefer that formulas be usable without this complication. -->

@@

<!-- ------------------------------ -->

<!-- ## Comparison dimensions -->

<!-- The tools for nonlinear estimation we consider have been built with different -->
<!-- goals and contexts. We will make our comparison considering the overlapping -->
<!-- headings summarized below. -->

<!-- ### Modeling versus minimizing a sum of squares -->

<!-- The goal of many statisticians and other workers is to estimate a model. The -->
<!-- minimization of a sum of squared residuals -- the nonlinear least squares problem -- -->
<!-- is the mechanism to estimate the model. Some tools focus on modeling,  -->
<!-- so attempt to provide measures of uncertainty and other information.  -->
<!-- The interface to modeling tools will generally try to streamline that task. By -->
<!-- contrast, nonlinear least squares tools are trying to solve the more limited -->
<!-- numerical task. -->

<!-- ### Programming language(s) -->

<!-- R tools are our interest, but they often call programs in other programming -->
<!-- languages. There are advantages in using a well-established program that has -->
<!-- been used and tested. Early versions of R were relatively sluggish in  -->
<!-- computational speed. The downside is that we must interface between two or more -->
<!-- programming languages, adding to the risk of errors. Changes to one or the  -->
<!-- other language or its library support may -->
<!-- force maintenance issues. Moreover, the pool of expertise with languages like C -->
<!-- and Fortran is diminishing. -->

<!-- ### Internal solver stabilization -->

<!-- Early nonlinear least squares programs generally used the Gauss-Newton method  -->
<!-- (@Hartley1961). This is often very effective and efficient, but for a quite large  -->
<!-- proportion of problems can fail to find the solution. Stabilization of the  -->
<!-- computations is relatively straightforward in principle, and overcomes most -->
<!-- of the failures, though sometimes at the cost of more calculations. -->

<!-- ### Jacobian calculation -->

<!-- The necessarily iterative solution of a nonlinear least squares problem  -->
<!-- to carry out nonlinear modeling requires that we can find smaller values -->
<!-- of a sum of squared residuals. This implies there is a "downhill" direction. -->
<!-- The gradient of the residual sum of squares is computed as its product -->
<!-- with the Jacobian matrix: the partial derivatives of the residual vector  -->
<!-- with respect to the modeling parameters. -->

<!-- Most researchers are challenged to program the differential calculus  -->
<!-- required to specify the Jacobian to a nonlinear estimation tool.  -->
<!-- The workaround is generally to use a numerical approximation, -->
<!-- but there is the possibility of using symbolic or automatic differentiation. -->
<!-- There are also particular models where workers have provided code for the -->
<!-- Jacobian. -->

<!-- ### Algorithm and termination choices -->

<!-- Iterative algorithms need to be stopped at some point, and how termination -->
<!-- is forced may alter results. Furthermore, even supposedly similar algorithms -->
<!-- may differ greatly in detail.  -->

<!-- ### Constraints on parameters -->

<!-- Reality imposes conditions on parameters. While general constraints can -->
<!-- be difficult to include, bounds on parameters can be accommodated by many  -->
<!-- solvers. For some reason, fixed parameters (masks) are not often considered -->
<!-- but are convenient when a parameter has a commonly accepted value that we -->
<!-- later wish to allow to be estimated from the data. -->

<!-- ### Special capabilities -->

<!-- Users often have particular problems that introduce other requirements. It is -->
<!-- common to weight observations where their reliability is thought to differ. -->
<!-- Some situations present a number of related situations where a combined model -->
<!-- with some common and some local parameters are present, and the local parameters -->
<!-- could be indexed. The common exponential decay and related **partially linear** -->
<!-- models (sometimes called **separable** models) is a difficult one that can  -->
<!-- benefit from special treatment.  -->

<!-- ### Output  -->

<!-- While nonlinear estimation aims to return a set of parameters for a -->
<!-- particular model, there is a variety of additional information which may be -->
<!-- provided. From the perspective of the nonlinear least squares problem, there -->
<!-- are measures of work and information on the quality of the solution, while -->
<!-- for statistical purposes measures of uncertainty in the parameters are helpful. -->

<!-- Some tools allow constraints to be imposed on the solution, and it is helpful -->
<!-- to know if these constraints are active in the proposed solution. -->

<!-- ### Documentation -->

<!-- We believe that the availability of good documentation with examples that users -->
<!-- can modify to deal with their problems is essential. The presence of vignettes,  -->
<!-- demos, and tests indicates a package is well-supported. -->

>>>>

---
title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
date: "2023-5-18"
abstract: >
  Our Google Summer of Code project "Improvements to `nls()`" considered the 
  features and limitations of the R function `nls()` with the aim 
  of improving and rationalizing R tools for nonlinear regression.
  The rich features of `nls()` are weakened by several deficiencies and
  inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
  Further considerations are the usability and  maintainability of
  the code base that provides the functionality `nls()` claims to offer.
  Various packages, including our `nlsr`, provide alternative capabilities.
  We consider the differences in goals, approaches, and features of
  different tools for nonlinear least squares modeling in R. Discussion of 
  these matters is relevant to improving R generally as well as its 
  nonlinear estimation tools.

draft: true
author:
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output: 
  rjtools::rjournal_web_article:
      self_contained: yes
      toc: no
bibliography: ./ComparingNLStools.bib
---

@@

### Principal messages

Our work was aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored `nls()` function.

For R **users**, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package `nlsr` or the example we give below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

For R **developers** we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features and 
efficiency of ongoing maintenance.

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ g = 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ 0.5 * H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ 0.5 * H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by NOT forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

As mentioned, the output of `nls()`, `minpack.lm::nlsLM()`, or
`gslnls::gsl_nls()` is an object of class "nls" which has a quite rich structure
described in the manual files or revealed by applying the `str()` function to the
result of `nls()`.  The complexity of this object is a challenge to users. Let us 
use for example `result <- snlm1` as the returned object from `nlsLM()` for the 
Logistic3S problem. . The `data` return element 
is an R symbol. To actually access the data from this element, we need to use 
the syntax:
```
eval(parse(text=result$data))
```
However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Bounds on parameters raise some interesting questions about how 
uncertainty in parameter estimates should be computed or reported. That is, the
traditional "standard errors" are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions and hence the interpretation of estimates of
uncertainty, whether by traditional approximations from the $J' J$ matrix or from methods
such as profile likelihood or bootstrap. At the time of writing, 
`nlsr::nlxb()` does not compute standard errors nor their derived statistics 
when bounds are active to avoid providing misleading information.

@@

@@

@@

### 2023-05-18 JNABsub20230518/ComparingNLStools.tex
May 18 fixes

@@

Let us consider there are \(n\) parameters and \(m\) residuals.
The loss function is
\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
The gradient of \(S(p)\) is
\[ 2 * J' r\]
where the Jacobian \(J\) is given by elements
\[ J_{i,j} = \partial r_i / \partial p_j \]
and the Hessian is defined by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
If we expand the Hessian for nonlinear least squares problems, we find
\[  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this expression.
What is generally called \textbf{Newton's method} for function minimization tries
to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
This leads to \textbf{Newton's equation}
\[ H \delta = -g \]
Given a set of parameters \(p\), we solve this equation for \(\delta\), adjust \(p\)
to \(p + \delta\) and iterate, hopefully to converge on a solution.
Applying this to a sum of squares problem gives
\[ H \delta = (J'J + D) \delta = - J' r\]
In this expression, only the elements of \(D\) have second partial derivatives.
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of \(r\), so he approximated
\[ H \approx J' J\]
by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
solve
\[ J' J \delta = - J' r\]
though we can avoid some loss of accuracy by NOT forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \]
by one of several matrix decomposition methods.

@@

As mentioned, the output of \texttt{nls()} is an object of class ``nls'' which has a quite rich structure
described in the manual file or revealed by applying the \texttt{str()} function to the
result of \texttt{nls()}.

The complexity of this object is a challenge to users. Let us use the
\texttt{result} as the returned object from \texttt{nls()}, \texttt{minpack.lm::nlsLM()} or,
\texttt{gslnls::gsl\_nls()}. For example, the \texttt{data} return element is an R symbol. To
actually access the data from this element, we need to use the syntax:

@@

Bounds on parameters raise some interesting questions about how
uncertainty in parameter estimates should be computed or reported. That is, the
traditional ``standard errors'' are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not compute standard
errors nor their derived statistics when bounds are active.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}, which we
have had to extract from the base R code and explicitly \texttt{source()} because it
is not exported to the working namespace,
allows us to compute residuals for particular coefficient sets.

@@

\begin{verbatim}
source("nlsModel.R")
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0
\end{verbatim}

@@

\bibliography{./ImproveNLS.bib}

>>>>

\hypertarget{principal-messages}{%
\subsubsection{Principal messages}\label{principal-messages}}

Our work was aimed at unifying nonlinear modeling functionality in R,
ideally in a refactored \texttt{nls()} function.

For R \textbf{users}, we would advise adapting working scripts, preferably those
with documentation and using recent tools. If there is a suspicion that there
may be ill-conditioning, package \texttt{nlsr} or the example we give below of how to
find singular values of the Jacobian allow these diagnostics to be calculated.

For R \textbf{developers} we invite and encourage discussion of the design choices, since these
have downstream implications for ease of use, adaptation to new features and
efficiency of ongoing maintenance.

@@

Let us consider there are \(n\) parameters and \(m\) residuals.
The loss function is
\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
The gradient of \(S(p)\) is
\[ g = 2 * J' r\]
where the Jacobian \(J\) is given by elements
\[ J_{i,j} = \partial r_i / \partial p_j \]
and the Hessian is defined by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
If we expand the Hessian for nonlinear least squares problems, we find
\[  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this expression.
What is generally called \textbf{Newton's method} for function minimization tries
to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
This leads to \textbf{Newton's equation}
\[ H \delta = -g \]
Given a set of parameters \(p\), we solve this equation for \(\delta\), adjust \(p\)
to \(p + \delta\) and iterate, hopefully to converge on a solution.
Applying this to a sum of squares problem gives
\[ 0.5 * H \delta = (J'J + D) \delta = - J' r\]
In this expression, only the elements of \(D\) have second partial derivatives.
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of \(r\), so he approximated
\[ 0.5 * H \approx J' J\]
by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
solve
\[ J' J \delta = - J' r\]
though we can avoid some loss of accuracy by NOT forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \]
by one of several matrix decomposition methods.

@@

As mentioned, the output of \texttt{nls()}, \texttt{minpack.lm::nlsLM()}, or
\texttt{gslnls::gsl\_nls()} is an object of class ``nls'' which has a quite rich structure
described in the manual files or revealed by applying the \texttt{str()} function to the
result of \texttt{nls()}. The complexity of this object is a challenge to users. Let us
use for example \texttt{result\ \textless{}-\ snlm1} as the returned object from \texttt{nlsLM()} for the
Logistic3S problem. . The \texttt{data} return element
is an R symbol. To actually access the data from this element, we need to use
the syntax:

@@

Bounds on parameters raise some interesting questions about how
uncertainty in parameter estimates should be computed or reported. That is, the
traditional ``standard errors'' are generally taken to imply symmetric intervals about
the point estimate in which the parameter may be expected to be found with some
probability under certain assumptions.
Bounds change those assumptions and hence the interpretation of estimates of
uncertainty, whether by traditional approximations from the \(J' J\) matrix or from methods
such as profile likelihood or bootstrap. At the time of writing,
\texttt{nlsr::nlxb()} does not compute standard errors nor their derived statistics
when bounds are active to avoid providing misleading information.

@@

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
allows us to compute residuals for particular coefficient sets.
We have had to extract from the base R code and include it via a code chunk
(not echoed here for space) because it is not exported to the working namespace,
We could also explicitly \texttt{source()} this code.

@@

\begin{verbatim}
## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
attr(rn0,"gradient")<-NULL; rn0
\end{verbatim}

@@

\bibliography{./ComparingNLStools.bib}
